The kinematics of the Local Group in a cosmological context

### 2013-03-07 paper/ms.tex
accepted version :)

\begin{abstract}
Recent observations constrained the tangential velocity of M31 with
respect to the Milky Way (MW) to be $v_{\rm M31,tan}<34.4$ \kms and
the radial velocity to be in the range $v_{\rm M31,rad}=-109\pm
4.4$\kms \citep{vanderMarel12}. In this study we use a large volume
high resolution N-body cosmological simulation (Bolshoi) together with three
constrained simulations to statistically study this kinematics in the
context of the $\Lambda$CDM. The comparison of the ensembles of
simulated pairs with the observed LG {\bf at the 1-$\sigma$ level in
  the uncertainties} has been done with respect to the
radial and tangential velocities, the reduced orbital energy
($e_{\rm tot}$), angular momentum ($l_{\rm orb}$) and the
dimensionless spin parameter, $\lambda$. Our main results are: (i) the
preferred radial and tangential velocities for pairs in $\Lambda$CDM
are $v_{\rm r}=-80\pm 20$\kms, $v_{\rm   t}=50\pm 10$\kms, (ii) pairs
around that region are $3$ to $13$ times more common than pairs within
the observational values, (iii) $15\%$ to $24\%$ of LG-like pairs in
$\Lambda$CDM have energy and angular momentum consistent with
observations while (iv) $9\%$ to $13\%$ of pairs in the same sample
show similar values in the inferred dimensionless spin parameter. It
follows that within current observational uncertainties the
quasi-conserved quantities that characterize the orbit of
the LG, i.e. $e_{\rm tot}$, $l_{\rm   orb}$ and $\lambda$, do not
challenge the standard $\Lambda$CDM model, but the model is in tension
with regard to the actual values of the radial and tangential
velocities. This might hint to a problem of the $\Lambda$CDM model to
reproduce the observed LG.

@@

\item Values in parenthesis correspond to vector
  components. $\sigma_{\bf x}$ represents the uncertainty on the
  components of vector ${\bf x}$. {\bf The uncertainties correspond to
    1-$\sigma$ values.}  
\item The values for the individual halo
masses are consistent with the priors used by \cite{vanderMarel12}.

@@

\begin{figure*}
\begin{center}
\includegraphics[keepaspectratio=true,width=0.42\textwidth]{fig_1a.eps}
\includegraphics[keepaspectratio=true,width=0.42\textwidth]{fig_1b.eps}
\includegraphics[keepaspectratio=true,width=0.42\textwidth]{fig_1c.eps}
\includegraphics[keepaspectratio=true,width=0.42\textwidth]{fig_1d.eps}
\caption{Upper panels: 2D histograms of the radial and tangential
  velocities for LG-like halo pairs in the Bolshoi simulation.   Lower
  panels: 2D histograms of the orbital angular momentum ($l_{\rm orb}$)
  and mechanical energy ($e_{\rm tot}$) per unit of reduced mass
  calculated considering the halos as point masses. {\bf Left (right) panels
    correspond to the full (reduced) sample}. The bar indicates
  the number of pairs in each cell. The locii for the high
  density regions in the $v_{\rm r}-v_{\rm t}$ plane are listed in
  Table \ref{table:2}.   
  The half ellipse in the $v_{\rm r}-v_{\rm t}$ plane corresponds to
  the {\bf 1-$\sigma$ uncertainties} in observations. Equivalently, the contour
  line in   the $e_{\rm tot}-l_{\rm orb}$ plane encloses $68\%$ of the
  Monte   Carlo generated points from
  the   observational values summarized in Table 1. The circles
  represent the positions of the pairs from the constrained simulations.
}
\label{fig:rt}
\end{center}

@@

\begin{table*}
\caption{Summary of the comparison of the observational results
  against $\Lambda$CDM.}
\begin{center}
\begin{tabular}{ccccc}\hline
Physical & (\%) Pairs consistent & (\%) Pairs consistent & (\%) Pairs
with highest & (\%) with highest\\ 
property & with observations (1-$\sigma$) & with
observations (1-$\sigma$)& likelihood in $\Lambda$CDM & likelihood in $\Lambda$CDM\\ 
 & (full sample) & (reduced sample) & (full sample) & (reduced sample)\\ \hline
$v_{\rm r}$-$v_{\rm t}$ & (0.4\%) 8/1923 & ($<$0.6\%)0/158 & (1\%)23/1923 & (8\%)13/158\\
$e_{\rm tot}$-$l_{\rm orb}$ & (15\%)298/1923 & (24\%)38/158 & - & -\\
$\log_{10}\lambda$ & (13\%)257/1923 & (9\%)15/158 & - & -\\
$r_{\rm t}=v_{\rm t}/v_{\rm r}$& (12\%)242/1923 & (8\%)13/158 & -& -\\\hline
\end{tabular}
\end{center}
\vspace{1mm}
Notes:\\
\begin{enumerate}
\item 
{\bf In columns 2 and 3 consistency is defined from 1-$\sigma$
  uncertainties for $v_{\rm r}$-$v_{\rm t}$, $e_{\rm tot}$-$l_{\rm
    orb}$ and $\log_{10}\lambda$. In the case of $r_{\rm t}$, is
  defined from the constrain $r_{\rm t}<0.32$ derived from the
  1$-\sigma$ observational uncertainties in the radial and tangential
  velocities.} 
\item
In columns 4 and 5 the number of pairs around the preferred region
in $v_{\rm r}-v_{\rm t}$ in $\Lambda$CDM (as shown in
Fig. \ref{fig:rt} and summarized in Table \ref{table:2}) are
calculated using the same observational 1-$\sigma$ absolute
uncertainty on $v_{\rm r}$ and $v_{\rm t}$.
\end{enumerate}
\label{table:3}
\end{table*}

@@

The number of pairs compatible with LG observations {\bf at the
  1-$\sigma$ level} are listed in
Table \ref{table:3} (columns 2 and 3). In the same table (columns 4
and 5) we summarize the number of pairs around the $\Lambda$CDM values
within the same range of absolute observational uncertainty 
(i.e. $\sigma_{\rm tan}=17$\kms and $\sigma_{\rm rad}=4$\kms.)

From these results we infer that the pairs around the preferred phase
space for $\Lambda$CDM are at least $13$ times more common than pairs
with the observed velocities for the LG. We highlight that
this is a lower bound given that in the reduced sample none of the
pairs is found in the interval allowed by observations. The high
eccentric orbit of the observed LG  constitutes an unlikely
configuration for the $\Lambda$CDM LG-like objects. This holds for
the full and the reduced sample of objects. {\bf These
  conclusions are valid at 1-$\sigma$ level in the observational
  uncertainties. An equivalence in the abundance between the pairs
  around the observational values and those around the
  preferred $\Lambda$CDM velocities is reached only at
  the 4-$\sigma$ level on the observational uncertainties.}

@@

Summarizing, we see a broad agreement in the total angular
momentum and energy and a marked difference with the precise balance
between today's radial and tangential velocities, {\bf i.e. head-on
  collisions of MW-like halos, as described by 1-$\sigma$
  uncertainties in observations, are not common in $\Lambda$CDM}.

Under the
approximation of conservation of the orbital angular momentum and
mechanical energy, this could only be explained if the initial
conditions for the formation of the Local Group are special in
comparison to the initial conditions of any other pair of dark matter
halos in the $\Lambda$CDM cosmology. {\bf Investigating this
  perspective, by a thorough characterization  of the small pair
  sample consistent with observations in terms   of halo properties
  such as concentration, spin and place in the cosmic web
  \citep{Forero-Romero09,Hoffman12} is  underway (Bustamante et al. in
  prep.)}.

{\bf The results presented in this Letter open a new window into the
  question of how unique, if at all, is the LG in a cosmological
  context. This will continue to be studied within the framework of
  the CLUES project.}

>>>>

\begin{abstract}
Recent observations constrained the tangential velocity of M31 with
respect to the Milky Way (MW) to be $v_{\rm M31,tan}<34.4$ \kms and
the radial velocity to be in the range $v_{\rm M31,rad}=-109\pm
4.4$\kms \citep{vanderMarel12}. In this study we use a large volume
high resolution N-body cosmological simulation (Bolshoi) together with three
constrained simulations to statistically study this kinematics in the
context of the $\Lambda$CDM. The comparison of the ensembles of
simulated pairs with the observed LG at the 1-$\sigma$ level in
  the uncertainties has been done with respect to the
radial and tangential velocities, the reduced orbital energy
($e_{\rm tot}$), angular momentum ($l_{\rm orb}$) and the
dimensionless spin parameter, $\lambda$. Our main results are: (i) the
preferred radial and tangential velocities for pairs in $\Lambda$CDM
are $v_{\rm r}=-80\pm 20$\kms, $v_{\rm   t}=50\pm 10$\kms, (ii) pairs
around that region are $3$ to $13$ times more common than pairs within
the observational values, (iii) $15\%$ to $24\%$ of LG-like pairs in
$\Lambda$CDM have energy and angular momentum consistent with
observations while (iv) $9\%$ to $13\%$ of pairs in the same sample
show similar values in the inferred dimensionless spin parameter. It
follows that within current observational uncertainties the
quasi-conserved quantities that characterize the orbit of
the LG, i.e. $e_{\rm tot}$, $l_{\rm   orb}$ and $\lambda$, do not
challenge the standard $\Lambda$CDM model, but the model is in tension
with regard to the actual values of the radial and tangential
velocities. This might hint to a problem of the $\Lambda$CDM model to
reproduce the observed LG.

@@

\item Values in parenthesis correspond to vector
  components. $\sigma_{\bf x}$ represents the uncertainty on the
  components of vector ${\bf x}$. The uncertainties correspond to
    1-$\sigma$ values.
\item The values for the individual halo
masses are consistent with the priors used by \cite{vanderMarel12}.

@@

\begin{figure*}
\begin{center}
\includegraphics[keepaspectratio=true,width=0.42\textwidth]{fig_1a.eps}
\includegraphics[keepaspectratio=true,width=0.42\textwidth]{fig_1b.eps}
\includegraphics[keepaspectratio=true,width=0.42\textwidth]{fig_1c.eps}
\includegraphics[keepaspectratio=true,width=0.42\textwidth]{fig_1d.eps}
\caption{Upper panels: 2D histograms of the radial and tangential
  velocities for LG-like halo pairs in the Bolshoi simulation.   Lower
  panels: 2D histograms of the orbital angular momentum ($l_{\rm orb}$)
  and mechanical energy ($e_{\rm tot}$) per unit of reduced mass
  calculated considering the halos as point masses. Left (right) panels
    correspond to the full (reduced) sample. The bar indicates
  the number of pairs in each cell. The locii for the high
  density regions in the $v_{\rm r}-v_{\rm t}$ plane are listed in
  Table \ref{table:2}.   
  The half ellipse in the $v_{\rm r}-v_{\rm t}$ plane corresponds to
  the 1-$\sigma$ uncertainties in observations. Equivalently, the contour
  line in   the $e_{\rm tot}-l_{\rm orb}$ plane encloses $68\%$ of the
  Monte   Carlo generated points from
  the   observational values summarized in Table 1. The circles
  represent the positions of the pairs from the constrained simulations.
}
\label{fig:rt}
\end{center}

@@

\begin{table*}
\caption{Summary of the comparison of the observational results
  against $\Lambda$CDM.}
\begin{center}
\begin{tabular}{ccccc}\hline
Physical & (\%) Pairs consistent & (\%) Pairs consistent & (\%) Pairs
with highest & (\%) with highest\\ 
property & with observations (1-$\sigma$) & with
observations (1-$\sigma$)& likelihood in $\Lambda$CDM & likelihood in $\Lambda$CDM\\ 
 & (full sample) & (reduced sample) & (full sample) & (reduced sample)\\ \hline
$v_{\rm r}$-$v_{\rm t}$ & (0.4\%) 8/1923 & ($<$0.6\%)0/158 & (1\%)23/1923 & (8\%)13/158\\
$e_{\rm tot}$-$l_{\rm orb}$ & (15\%)298/1923 & (24\%)38/158 & - & -\\
$\log_{10}\lambda$ & (13\%)257/1923 & (9\%)15/158 & - & -\\
$r_{\rm t}=v_{\rm t}/v_{\rm r}$& (12\%)242/1923 & (8\%)13/158 & -& -\\\hline
\end{tabular}
\end{center}
\vspace{1mm}
Notes:\\
\begin{enumerate}
\item 
In columns 2 and 3 consistency is defined from 1-$\sigma$
  uncertainties for $v_{\rm r}$-$v_{\rm t}$, $e_{\rm tot}$-$l_{\rm
    orb}$ and $\log_{10}\lambda$. In the case of $r_{\rm t}$, is
  defined from the constrain $r_{\rm t}<0.32$ derived from the
  1$-\sigma$ observational uncertainties in the radial and tangential
  velocities.
\item
In columns 4 and 5 the number of pairs around the preferred region
in $v_{\rm r}-v_{\rm t}$ in $\Lambda$CDM (as shown in
Fig. \ref{fig:rt} and summarized in Table \ref{table:2}) are
calculated using the same observational 1-$\sigma$ absolute
uncertainty on $v_{\rm r}$ and $v_{\rm t}$.
\end{enumerate}
\label{table:3}
\end{table*}

@@

The number of pairs compatible with LG observations  at the
  1-$\sigma$ level are listed in
Table \ref{table:3} (columns 2 and 3). In the same table (columns 4
and 5) we summarize the number of pairs around the $\Lambda$CDM values
within the same range of absolute observational uncertainty 
(i.e. $\sigma_{\rm tan}=17$\kms and $\sigma_{\rm rad}=4$\kms.)

From these results we infer that the pairs around the preferred phase
space for $\Lambda$CDM are at least $13$ times more common than pairs
with the observed velocities for the LG. We highlight that
this is a lower bound given that in the reduced sample none of the
pairs is found in the interval allowed by observations. The high
eccentric orbit of the observed LG  constitutes an unlikely
configuration for the $\Lambda$CDM LG-like objects. This holds for
the full and the reduced sample of objects. These
  conclusions are valid at 1-$\sigma$ level in the observational
  uncertainties. An equivalence in the abundance between the pairs
  around the observational values and those around the
  preferred $\Lambda$CDM velocities is reached only at
  the 4-$\sigma$ level on the observational uncertainties.

@@

Summarizing, we see a broad agreement in the total angular
momentum and energy and a marked difference with the precise balance
between today's radial and tangential velocities, i.e. head-on
  collisions of MW-like halos, as described by 1-$\sigma$
  uncertainties in observations, are not common in $\Lambda$CDM.

Under the
approximation of conservation of the orbital angular momentum and
mechanical energy, this could only be explained if the initial
conditions for the formation of the Local Group are special in
comparison to the initial conditions of any other pair of dark matter
halos in the $\Lambda$CDM cosmology. Investigating this
  perspective, by a thorough characterization  of the small pair
  sample consistent with observations in terms   of halo properties
  such as concentration, spin and place in the cosmic web
  \citep{Forero-Romero09,Hoffman12} is  underway (Bustamante et al. in
  prep.).

The results presented in this Letter open a new window into the
  question of how unique, if at all, is the LG in a cosmological
  context. This will continue to be studied within the framework of
  the CLUES project.
